---
title: "Day 1: Session Setup"
author: "Niklas Schandry"
date: "April 2020"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 16
    highlight: haddock
    theme: cosmo
    toc: yes
    toc_float: yes
---

Below installs first the "renv" (R environments) package, which then prepares all packages that are required for this project. This is done via renv::restore().
This is cool and will save you some time you would otherwise spend manually installing a ton of packages and will hopefully avoid packages conflicts.
However, we need bioconductor to install some packages.

```{r}
install.packages("BiocManager")
BiocManager::install(version = "3.10")
```

Then we install renv and restore the environment.

For this server, I have cached the packages into a shared directory, so it will take some minutes.
If you were to install all of these packages, it would probably take 1-2 hours.

```{r}
install.packages("renv")
renv::restore()
```

Run the code above and let me know if you encounter errors or warnings.



