---
title: "Workload Metagenomics course"
author: "Niklas Schandry"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    highlight: haddock
    theme: cosmo
    toc: yes
    toc_float: yes
---

```{r echo = F, message=F}
library(tidyverse)
seminar <- tibble::tibble("Task" = c("Collecting information on the topic",
                                     "Preparations of presentation (own)",
                                     "Practicing presentation (own)",
                                     "Refinement of presentation (own)",
                                 "Feeback & reflection on own presentation (own, FT)",
                                 "Participation (10x Presentations; FT)",
                                 "Reflection of presentation (9x)",
                                 "Paper reading (10x)",
                                 "Paper discussion (10x; FT)",
                                 "Reflection of manuscript discussion (10x)"),
                      "Time (h)" = c(4, 5, 5, 5, 2, 5, 5, 40, 9, 10),
                      "Type" = c("Seminar / Presentation", 
                                 "Seminar / Presentation", 
                                 "Seminar / Presentation", 
                                 "Seminar / Presentation", 
                                 "Seminar / Presentation",
                                 "Seminar / Presentation",
                                 "Seminar / Presentation",
                                 "Seminar / Paper Discussion",
                                 "Seminar / Paper Discussion", 
                                 "Seminar / Paper Discussion"))

course <- tibble::tibble("Task" = c("Working through daily protocol (12 days)", "Assignments on Moodle (12 days)", "Daily reflection (12 days)", "Final Discussion and report"),
                     "Time (h)" = c(60, 12, 6, 12),
                     "Type" = "Course",
                     )
                      
```

# About this document

This document tabulates the ECTS workload of the Seminar + Course ('Übung') _19442: R-Course and Seminar: Metagenomics + Community analysis_.

# General Info

One ECTS equals 30 hours of 'student' work. 19442 consists of a Seminar (2 ECTS) and a practical-course (3 ECTS).

# Tabulations

## Grand total

Total hours:
```{r, echo = F}
rbind(seminar, course) %>% 
  group_by(Type) %>%
  gt::gt() %>% 
   gt::grand_summary_rows(
   columns = "Time (h)",
   fns = list(
     Total = ~sum(., na.rm = TRUE))
   ) %>% 
  gt::tab_header(
   title = gt::md("Breakdown of hours"),
   subtitle = gt::md("Metagenomics course (19442)")) %>% 
   gt::tab_footnote(
    footnote = "FT = Face Time",
    locations = gt::cells_column_labels(columns = vars(Task)))
```
With 1 ECTS equalling 30 hours, this is worth 5 ECTS.

## Seminar

The seminar is divided into presentation, and paper discussion. There will be 10 presentations, that generally introduce a topic, and 10 seminars, where a paper is discussed. The presentations and discussion of papers may take place on different days.

```{r, echo = F}
seminar %>%
  group_by(Type) %>%
  gt::gt() %>% 
    gt::grand_summary_rows(
    columns = "Time (h)",
    fns = list(
      Total = ~sum(., na.rm = TRUE))
    ) %>%
  gt::tab_header(
    title = gt::md("Breakdown of hours for the Seminar"),
    subtitle = gt::md("Metagenomics course (19442)")
  )  %>% 
  gt::tab_footnote(
    footnote = "FT = Face Time",
    locations = gt::cells_column_labels(columns = vars(Task)))
```

With 1 ECTS equalling 30 hours, this is worth 2 ECTS.

## Course ETCS

Tabulation of hours for the practical:

```{r, echo = F}
course %>%
  group_by(Type) %>%
  gt::gt() %>% 
    gt::grand_summary_rows(
    columns = "Time (h)",
    fns = list(
      Total = ~sum(., na.rm = TRUE))
    ) %>%
  gt::tab_header(
    title = gt::md("Breakdown of hours for the Course"),
    subtitle = gt::md("Metagenomics course (19442)")
  )
```

With 1 ECTS equalling 30 hours, this is worth 3 ECTS.